Application for Rethinking Foundations of Physics 2017

### 2017-01-11 Rethinking Foundations of Physics 2017/application.tex
have fun Yu-Tsung

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage[ backend=biber]{biblatex}
%\addbibresource{../proposal/prop.bib}
\addbibresource{C:/Users/Yu-Tsung/workspace/Contextuality-and-Nonlocality-in-Discrete-Quantum-Theory/proposal/prop.bib}
\ExecuteBibliographyOptions{sorting=none,maxbibnames=5,doi=false,isbn=false,url=false}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{mathrsfs}
\usepackage{bbold}
\usepackage{wesa}

@@

It is widely believed that the Turing machine is equivalent to any
reasonable computational models. Yet, for centuries, the mathematical
formalization of physical processes has been founded on the infinitely
precise real or complex numbers~\cite{Ziegler2007,weihrauch2012computable,blum2012complexity}
which might provide such processes more computational power. For example,
almost every description of quantum mechanics, quantum computation,
or quantum experiments refers to entities such as $\rme$, $\pi$,
$\sqrt{2}$, etc (see, e.g., \cite{Redhead1987-REDINA,544199,Mermin2007}).
Whether these models based on real numbers faithfully describe the
nature or just an approximation is still an open question. If they
are faithful, how the nature store and compute these real numbers
to produce our experimental results would be the next natural question.
There are at least three possibilities:
\begin{itemize}
\item For any computable number~\cite{Turing_1937}, the nature only store
the code of the Turing machine which computes the given number, and
manipulate these codes directly during the computation~\cite{Ziegler2007,weihrauch2012computable}.
\item The nature stores real numbers on infinitely long Turing machine tapes,
and handle these numbers by a Turing machine~\cite{Ziegler2007,weihrauch2012computable}.
In another word, the nature stores real numbers as a whole, but compute
them piecewisely.
\item The nature stores and computes real numbers as a whole~\cite{Ziegler2007,blum2012complexity}.
\end{itemize}
All these possibilities are not very desirable because they suggest
the nature might have more computational power than a Turing machine.
On one hand, if the nature only stores computable numbers by their
Turing machine codes, the nature might be able to decide whether two
Turing machines output exactly the same numbers, since this is used
to decide whether an observable is degenerated or not. However, testing
whether two Turing machines are output-equivalent is undecidable by
Rice's theorem~\cite{Rice_1953,Ziegler2007,weihrauch2012computable},
i.e., the nature has more computational power than a Turing machine.
On the other hand, if the nature stores real numbers as a whole, any
particular real number could serve as an oracle and give the nature
more computational power than a Turing machine. For example, maybe
the mass of the electron in some representation is the halting oracle
for some representation of Turing machines. Then, we could solve the
halting problem by just measuring the mass of the electron. This kind
of possibility has been online discussed by Peter Shor and others~\cite{PhysicsStackExchange16889}.

\section{Proposed Question}

@@

@@

\begin{comment}
 \bibliographystyle{plain}
\bibliography{../proposal/prop}

\end{comment}
\printbibliography 
\end{document}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage[ backend=biber]{biblatex}
%\addbibresource{../proposal/prop.bib}
%%\addbibresource{C:/Users/Yu-Tsung/workspace/Contextuality-and-Nonlocality-in-Discrete-Quantum-Theory/proposal/prop.bib}
%% \addbibresource{"../proposal/prop.bib"}
\addbibresource{prop.bib}
\ExecuteBibliographyOptions{sorting=none,maxbibnames=5,doi=false,isbn=false,url=false}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{mathrsfs}
\usepackage{bbold}
\usepackage{wesa}

@@

Physical processes are ultimately bounded in space, time, energy, and
other resources. Yet, for centuries, the mathematical formalization of
such processes has been founded on the infinitely precise real or
complex numbers. Indeed, physics textbooks casually refer to entities
such as $e$, $\pi$, $\sqrt{2}$, etc. From a computational perspective,
such numbers do not exist in their entirety ``for free.''  For
example, the state of the art algorithms for computing the $n$th
binary digit of~$\pi$ require on the order of $O(n\log^{O(1)}(n))$
operations~\cite{journals/moc/BaileyBP97}. In other words, simply
referring to the $n$th digit of $\pi$ requires more and more resources
as $n$ gets larger. Taking such resource bounds into consideration is
what founded computer science as a discipline and is crucial for
understanding the very nature of computation and, following
Feynman~\cite{Feynman1982Simulating}, Landauer~\cite{Landauer1996188},
and others, for understanding the very nature of physical processes.

\begin{quote}
  We therefore believe that the foundations of physics must be
  revisited from a computational perspective that only allows
  finitely-communicable evidence \emph{and} accounts for the resources
  needed to calculate, store, and communicate such evidence.
\end{quote}

\noindent This marriage of quantum mechanics and computer science first
envisioned and popularized by Feynman has created an awkward, but
opportune, moment. The embarrassing dilemma was concisely described by
Aaronson as follows~\cite{AaronsonArkhipov2011,BFRDARW2013}.  Consider
the following three statements:
\begin{enumerate}
\item[(A)] Textbook quantum mechanics is correct.
\item[(B)] There does not exist an efficient classical factoring
  algorithm.
\item[(C)] The extended Church-Turing thesis --- that probabilistic
  Turing machines can efficiently simulate any physically realizable
  model of computation --- is
  correct~\cite{BernsteinVazirani1997,Kaye2007}.
\end{enumerate}
There is overwhelming evidence to support each of these
statements. The theoretical framework of quantum mechanics (A) has
withstood decades of experimental confirmation. Entire industries are
founded on the assumption (B) that algorithms like RSA are secure and
they also have withstood years of attempted
attacks~\cite{boneh1999twenty,wiki:RSAFactoring}. Finally the entire
field of complexity theory in computer science which has also
withstood years of field testing rests, in essence, on assumption
(C)~\cite{Aaronson2005,Piccinini2015}.  And yet at least one of these
three statements \emph{must be false}! Indeed if textbook quantum
mechanics is false we concede (A). If it is correct Shor's efficient
factoring algorithm is realizable. If there is a corresponding
efficient classical factoring algorithm then we concede (B). If not
then we concede~(C).

It is unlikely that there will be a simple a resolution to this
awkward situation. It is more likely that the resolution will emerge
from deep and careful analyses of the foundations of each field. As
the world-renowned logician Girard states, the foundations of physics
and the foundations of logic and computation appear strongly
correlated:
\begin{quote}
  In other terms, what is so good in logic that quantum physics should
  obey? Can't we imagine that our conceptions about logic are wrong,
  so wrong that they are unable to cope with the quantum miracle?
  [\ldots] Instead of teaching logic to nature, it is more reasonable
  to learn from her. Instead of interpreting quantum into logic, we
  shall interpret logic into quantum~\cite{MSC:1444836}.
\end{quote}

%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Proposed Question}

The meaning of quantum probability

The reality of the quantum state

--- next section

The meaning of quantum probability: probability is approximate;
intervals; Born; no state when approximation is bad.

quantum state: Bayesian perspective; probability is subjective; hence
state is subjective; given previous results that show that state does
not exist when you have poor approximations; ?

@@

In
previous work, we examined the Hilbert space mathematical framework of
quantum mechanics and teased apart its various aspects that directly
rely on the infinite-precision real or complex numbers. We note two
remarkable earlier results and an open question:
\begin{itemize}
\item UNIQUE-SAT: Schumacher and
  Westmoreland~\cite{Schumacher2012-SCHMQT} argue that much of the
  structure of traditional quantum mechanics is maintained in the
  presence of finite fields. In particular, they establish that the
  quantum theory based on the finite field of booleans retains the
  following characteristics of quantum mechanics: the notions of
  superposition, interference, entanglement, and mixed states of
  quantum systems; the time evolution of quantum systems using
  invertible linear operators; the complementarity of incompatible
  observables; the exclusion of local hidden variable theories and the
  impossibility of cloning quantum states; and the presence of natural
  counterparts of quantum information protocols such as superdense
  coding and teleportation. However we have argued that this theory is
  not physically realizable as it allows the solution of
  computationally difficult problems like
  UNIQUE-SAT~\cite{Valiant198685} in constant time~\cite{usat}.

\item Deutsch-Jozsa: In a follow-up paper~\cite{DQT2014}, we proposed
  a variant of quantum mechanics in which the underlying field of
  complex numbers is replaced by special finite fields with enough
  structure to model Hermitian products that \emph{locally} behave
  like inner products. The trick is to recognize that, as the
  characteristic of the finite field increases, larger and larger
  regions of the finite field become totally ordered, thereby
  restoring, with some restrictions, all the usual properties of the
  inner product needed for conventional quantum mechanics. Using this
  theory to analyze the Deutsch-Jozsa algorithm reveals that the
  characteristic of the finite field plays an essential role: as the
  size of the finite field grows, so does the numerical range of the
  intermediate and final results of the algorithms being implemented.
  The theory naturally recovers a deterministic measure of the
  intrinsic resources required for a given level of complexity; this
  measure is normally completely hidden in computations with real
  numbers, and explicitly exposing it is one of the significant
  achievements of our discrete field analysis of quantum computation.
  This solves the conundrum that the conventional Deutsch-Jozsa
  algorithm~\cite{DeutschJozsa1992,Jaeger2007}
  mysteriously continues to work for larger and larger input
  functions without any apparent increase in resources.  Our analysis
  of this problem reveals that as the size of the input increases, it
  is necessary to increase the size of the finite field and hence the
  size of the underlying available numeric coefficients.  This
  observation does not fully explain the power of quantum computing
  over classical computing, but at least it explains that some of the
  power of quantum computing depends on increasingly larger precision
  in the underlying field of numbers.

\item An important special case of a generalized measurement is the
  symmetric, informationally complete, positive operator valued
  measure (SIC-POVM), which has applications in quantum
  cryptography. This problem has been formulated in the context of
  standard quantum theory where fields are complex numbers.  Although
  a rigorous proof for the existence of SIC-POVMs for arbitrary
  Hilbert space dimensions is still missing, it would be illuminating
  to look for a formulation of the same problem in a discrete quantum
  theory defined in terms of finite Galois fields.
\end{itemize}

A long term, ultimate, goal of this research proposal would be to
combine the various pieces into a coherent computational model of
quantum mechanics that keeps explicit track of resources needed to
represent the interactive system representing the quantum state, its
evolution, the observables, and the probabilities. Such a
computational model of quantum mechanics would encompass classical
(reversible) computation and potentially revolutionize quantum
thinking with finite resources. The explicit manipulation of resources
enables one of key idea that emerged in our previous work: the
resources used to represent the quantum system are in principle
unrelated to the resources available to the various observers. We
conjecture that this asymmetry may prove key to resolving some of the
quantum paradoxes.  The nature of quantum mechanics realized in terms
of discrete fields and finitely-communicable probability estimates is
amenable to direct computational exploration for reasonable field
sizes and observational accuracy. We have established a set of utility
libraries in Mathematica that supports a wide variety of such
numerical experiments. Concurrently to this effort, we are also
building on our previous work~\cite
{VDS2009,AGVS2005,VAS2006,finiteQC,Sabry:2003:MQC:871895.871900}
to build semantically well-founded implementations of the
computational models investigated in this proposal.

\paragraph*{Research Questions.} The aim of these implementations
include:
\begin{itemize}
\item Provide an experimental testbed to simulate physical phenomena
  (e.g.~\cite {OrtizEtAl2001,SommEtAl2002}) that complements the
  Mathematica libraries in which quantum algorithms can be simulated;
\item Establish connections between quantum programming concepts and
  classical programming concepts. We plan to make our software
  available for educational and research purposes; and 
\item Analyze the computational complexity of quantum theories and
  simulations based on discrete fields. What other (apart from
  UNIQUE-SAT) supernatural algorithms can be implemented in
  computational models based on discrete fields?
\item Determine, through numerical computations, solutions for
  SIC-POVM for small vector space dimensions.
\end{itemize}

propose to revisit quantum mechanics, quantum information, quantum
computation, and the foundations of physics from this resource-aware
perspective.

Soon after the discovery of quantum effects in the early 1900s, it
became evident that the quantum world requires a fresh worldview, not
merely an extension of the repertoire of entities and constructions
which explain classical physics. In the past thirty years, however,
the first quantum revolution mostly relied on just that: extensions of
classical logics, classical models of computation, and classical
analysis of algorithms. Fundamentally, the foundations of logic and
computing still embrace, to a large extent, the notion that the truth
of a statement is ``absolute'' and independent of any reasoning,
understanding, or action, i.e., logical statements are viewed as
abstract entities for which it makes no sense to talk about
entanglement with an observer: their truth or falsity is absolute and
does not depend on the notion of observation. This absolute view of
truth and the use of classical logic for reasoning about this absolute
truth have both been critiqued for centuries by philosophers and other
scientists but it is the advent of quantum science that really
devastates this worldview of classical determinism. Indeed, quantum
science suggests that the properties of physical systems can only be
learned by observers interacting with the system, and that this
interaction may entangle the observer with the system and disturb the
properties of interest. This suggests that the possibility structure
of our universe is not a fixed, Boolean structure, as we supposed
classically, but is in fact a non-Boolean structure that changes
dynamically~\cite{bub2006}.

We believe that the foundations of physics should be revisited from a
logical and computational approach

More directly, world-renowned logician Girard states:
Our vision for this revised computational and logic theory builds upon
programming abstractions which assume evolving, incomplete, and
inconsistent data, and principles for reasoning about programs which
are based on finitely communicable evidence between subsystems or
systems and observers, rather than absolute truth. The most natural
way to achieve this shift in our ``computational thinking'' and to
consolidate many of the new promising trends in computational,
informational, and logical theory is to design new programming
models. For it is indeed a widely accepted thesis that programming
languages shape thought processes and design
solutions~\cite{iverson80,johansen-berg01}. Our approach therefore is
to develop new programming models based on both logical foundations
and pragmatic considerations, to connect them to current
computational, informational, and logical theory, and to make
compelling arguments for the models by using them to write several
novel and challenging applications.

There is significant IU expertise in logic foundations, including
unconventional logics such paraconsistent logics, relevance logic,
multivalued logic with Bayesian inference, situation theory, and the
Dempster-Shafer theory of belief functions and plausible reasoning,
which can handle incomplete, evolving, and inconsistent information in
a controlled and discriminating way. There is also significant IU
expertise in the design and implementation of programming models, type
theories, and semantics that encompass both the classical and quantum
worlds. What IU lacks in expertise is an expert that bridges the
programming models with actual quantum algorithms: this is a new breed
of theoreticians with a strong background in both physics, complexity
theory, and cryptography.

Our initial results in that domain showed how subtle the
issues can be~\cite{usat,geometry2013,DQT2014}: a straightforward
replacement of the complex numbers by a finite field yields a variant
of quantum mechanics in which computationally hard problems like
UNIQUE-SAT (which decides whether a given Boolean formula is
unsatisfiable or has exactly one satisfying assignment) can be
deterministically solved in constant time. To eliminate such
unrealistic theories requires delicate analysis of the structure of
the Hilbert space, the process of observation, and the notion of
probability teasing apart their reliance on the infinitely precise
real numbers~\cite{geometry2013,DQT2014}. In this proposal, our aim is
to shift focus from the infinitely-specified but not directly
observable quantum states, to observable measurable properties of
quantum systems and their probabilities. Furthermore, we insist that
our theories of measurement and probability only refer to finitely
communicable evidence within feasible computational bounds. It follows
that states, observations, and probabilities all become ``fuzzy'',
i.e., specified by intervals of confidence that can only increase in
precision if the available resources increase proportionally. Our
notion of ``fuzzy quantum mechanics'' is related to existing
work~\cite{GranikCaulfield1996,Pykacz2013,SNL2009,Gudder2005,aerts1993physical}
but, as will be explained in more detail, is distinguished by its
unique computational character.

We will begin by reviewing existing work that recasts classical
probability spaces in a resource-aware setting and move to our
proposal which, briefly speaking, aims at recasting quantum
probability and hence quantum measurement to a corresponding
resource-aware setting.

In particular, the objectives to be achieved include:
\begin{itemize}
\item A computational framework for quantum measurement where quantum
  systems are described in terms of interactive processes. The
  framework takes care of the resources required (including the ones
  related to the observer) to determine a given property of the
  measured system.
\item Develop a measurement framework based on quantum interval-valued
  probabilities measures.
\item Assess the validity of fundamental theorems of quantum
  mechanics, such as Gleason, Bell, and Kochen-Specker, in a quantum
  formulation stripped away of infinite resources. We will aim to
  provide a definite answer to the Meyer-Mermin
  debate~\cite{PhysRevLett.83.3751, Mermin1999, BarrettKent2004} on
  the impact of finite precision measurements on the relevance of the
  Kochen-Specker
  theorem~\cite{kochenspecker1967,Redhead1987-REDINA,peres1995quantum,Jaeger2007}.
\item Study the computational complexity of quantum theories and
  simulations based on discrete (computable) fields.
\item Study the problem of finding a complete set of mutually unbiased
  bases
  (SIC-POVM)~\cite{1751-8121-45-8-085306,1751-8121-47-33-335302,Caves2002}
  over finite Galois fields.
\end{itemize}
We argue that no matter what the results are their impact will be
strong. Positive results that formulate a computable, with clear
resource bounds, theory of quantum measurement and quantum probability
would be essential for understanding and realizing quantum
computation. Any negative results would redirect research that aims at
realizing quantum computation to other approaches. At an intellectual
level, the results have the potential to solve or clarify several
paradoxes in quantum mechanics, quantum information, computability,
complexity theory, and perhaps quantum gravity.

For centuries, the mathematical formalization of physical processes
has been founded on the infinitely precise real or complex
numbers~\cite{Ziegler2007,weihrauch2012computable,blum2012complexity}.
For example, almost every description of quantum mechanics, quantum
computation, or quantum experiments refers to entities such as $\rme$,
$\pi$, $\sqrt{2}$, etc (see, e.g.,
\cite{Redhead1987-REDINA,544199,Mermin2007}).  Whether these models
based on real numbers faithfully describe the nature or just an
approximation is still an open question. If they are faithful, how the
nature store and compute these real numbers to produce our
experimental results would be the next natural question.  There are at
least three possibilities:
\begin{itemize}
\item For any computable number~\cite{Turing_1937}, the nature only store
the code of the Turing machine which computes the given number, and
manipulate these codes directly during the computation~\cite{Ziegler2007,weihrauch2012computable}.
\item The nature stores real numbers on infinitely long Turing machine tapes,
and handle these numbers by a Turing machine~\cite{Ziegler2007,weihrauch2012computable}.
In another word, the nature stores real numbers as a whole, but compute
them piecewisely.
\item The nature stores and computes real numbers as a whole~\cite{Ziegler2007,blum2012complexity}.
\end{itemize}
All these possibilities are not very desirable because they suggest
the nature might have more computational power than a Turing machine.
On one hand, if the nature only stores computable numbers by their
Turing machine codes, the nature might be able to decide whether two
Turing machines output exactly the same numbers, since this is used
to decide whether an observable is degenerated or not. However, testing
whether two Turing machines are output-equivalent is undecidable by
Rice's theorem~\cite{Rice_1953,Ziegler2007,weihrauch2012computable},
i.e., the nature has more computational power than a Turing machine.
On the other hand, if the nature stores real numbers as a whole, any
particular real number could serve as an oracle and give the nature
more computational power than a Turing machine. For example, maybe
the mass of the electron in some representation is the halting oracle
for some representation of Turing machines. Then, we could solve the
halting problem by just measuring the mass of the electron. This kind
of possibility has been online discussed by Peter Shor and others~\cite{PhysicsStackExchange16889}.

@@

\begin{comment}
\bibliographystyle{plain}
\bibliography{prop}
\end{comment}
\printbibliography 
\end{document}
